A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-08-17 arxiv/conclusions.tex
pf

In this work, we analyzed the bias and variance of multi-task learning versus single-task learning.
We provided tight concentration bounds for the bias and the variance.
Based on these bounds, we analyzed the impact of three properties, including task similarity, sample size, and covariate shift on the bias and variance, to derive conditions for transfer.
%Each of these results (based on each of the 3 properties) gives a different algorithmic/practical insight.
We validated our theoretical results.
Based on the theory, we proposed to train multi-task models by incrementally adding labeled data and showed encouraging results inspired by our theory.
We describe several open questions for future work.
First, our bound on the bias term (cf. Lemma \ref{lem_cov_derivative}) involves an error term that scales down with $\rho_1$.
Tightening this error bound might cover the unexplained observations in Figure \ref{fig_model_shift_phasetrans}.
Second, it would be interesting to extend our results to non-linear settings.
We remark that this likely requires addressing significant technical challenges  to deal with non-linearity.

>>>>

In this work, we analyzed the bias and variance of multi-task learning versus single-task learning.
We provided almost tight concentration bounds on the bias and variance.
Based on these bounds, we analyzed the impact of three properties, including task similarity, sample size, and covariate shift, on the bias and variance, based on which we derived conditions for transfer.
%Each of these results (based on each of the 3 properties) gives a different algorithmic/practical insight.
We validated our theoretical results.
Based on the theory, we proposed to train multi-task models by incrementally adding labeled data, and showed encouraging results inspired by our theory.
We describe several open questions for future work.
First, our bound on the bias term (cf. Lemma \ref{thm_model_shift}) involves an error term that scales down with $\rho_1$.
Tightening this error bound might cover the unexplained observations in Figure \ref{fig_model_shift_phasetrans}.
Second, it would be interesting to extend our results to non-linear settings.
We remark that this likely requires addressing significant technical challenges  to deal with non-linearity.

### 2020-08-17 arxiv/proof_main.tex
pf

\begin{proof}[Proof of Theorem \ref{thm_many_tasks}]
In this setting, we need to study the following loss function:
\begin{align}
	f(B; W_1, \dots, W_t) = \sum_{i=1}^t \bignorm{X B W_i - Y_i}^2,\label{eq_mtl_same_cov}
\end{align}
for $W_1, W_2, \dots, W_t \in \R^r$. Then it is not hard to derive a closed form solution for $B$ as
	\begin{align*}
		\hat{B}(W_1, \dots, W_t) &= (X^{\top}X)^{-1} X^{\top} \bigbrace{\sum_{i=1}^t Y_i W_i^{\top}} (\cal W  \cal W^{\top})^{-1} \\
		&= (B^\star \cal W ^{\top}) (\cal W \cal W ^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cal W \cal W^{\top})^{-1},
	\end{align*}
	where $\cal W \in\real^{r\times t}$ is a matrix defined as $\cal W=[W_1, W_2, \dots, W_t]$.
	%Now we switch $\hat{B}$ back into equation \eqref{eq_mtl_same_cov} to
Then as in \eqref{approxvalid}, we pick $N$ independent samples of the training set for each task with $N\ge n^{1-\e_0}$, and use the concentration result, Lemma \ref{largedeviation}, to get the validation loss as {\cor{start here}}
\be\label{eq_multival}g(\cal W)=  N\left[\val(\cal W) + t  \sigma^2 \right]\cdot \left( 1+\OO(p^{-(1-\e_0)/2+\e})\right).\ee
Here $\val(\cal W)$ is defined as
%it remains to consider minimizing the validation loss
	$$\val( \cal W):=\exarg{\varepsilon_j, \forall 1\le j\le t}{ \sum_{i=1}^t \bignorm{\Sigma^{1/2}( \hat B W_i - \beta_i)}^2} =  \delta_{\bias}(\cal W) + \delta_{\vari}(\cal W),$$
where the model shift bias term $\delta_{\bias}(\cal W) $ is given by
	\begin{align*}
		\delta_{\bias}(\cal W) :=\sum_{i=1}^t  \bignorm{\Sigma^{1/2}\bigbrace{(B^\star \cal W^\top) (\cal W\cal W^{\top})^{-1} W_i - \beta_i}}^2,
	\end{align*}
	and the variance term $\delta_{\vari}(\cal W)$ can be calculated as
	\begin{align*}
		\delta_{\vari}(\cal W):= \sigma^2 \cdot \bigtr{\Sigma (X^{\top}X)^{-1}}.
	\end{align*}
It suffices to minimize $\delta_{\bias}(\cal W)$ over $\cal W$, since both $t \sigma^2$ and $\delta_{\vari}$ do not depend on $\cal W$.

We denote $Q := \cal W^{\top} (\cal W\cal W^{\top})^{-1} \cal W \in\real^{k\times k}$, whose $(i,j)$-th entry is equal to $W_i^{\top} (\cal W\cal W^{\top})^{-1} W_j$.
	%Let $B^{\star} = [\beta_1, \beta_2, \dots, \beta_k] \in\real^{p \times k}$ denote the true model parameters.
	Now we can write $\delta_{\bias}(\cal W)$ succinctly as
	\begin{align*}
		\delta_{\bias}(\cal W) = \bignormFro{\Sigma^{1/2}B^{\star}  \bigbrace{Q -\id}}^2 .
	\end{align*}
	From this equation we can solve the minimizer optimally as $Q_0\equiv  \cal W_0^{\top} (\cal W_0\cal W_0^{\top})^{-1} \cal W_0=U_{r}U_r^{\top}$. On the other hand, let $\hat {\cal W}$ be the minimizer of $g$, and denote $\hat Q:= \hat{\cal W}^{\top} (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat{\cal W} $. We claim that $\hat Q$ satisfies
	\be\label{Q-Q}\|Q_0^{-1}\hat Q - \id\| = \oo(1) \quad \text{w.h.p.}\ee
	In fact, if \eqref{Q-Q} does not hold, then using the condition $\lambda_{\min}((B^{\star})^\top\Sigma B^{\star})\gtrsim \sigma^2$ and that $\delta_{\vari}(\cal W)=\OO(\sigma^2)$ by \eqref{eq_isometric}, we obtain that
	$$   \val(\hat {\cal W}) + t  \sigma^2 > (\val( {\cal W}_0) + t \sigma^2 )\cdot (1+\oo(1)) \ \Rightarrow \ g( \hat{\cal W})>g( {\cal W}_0),$$
	that is, $\hat {\cal W}$ is not a minimizer. This leads to a contradiction.

	In sum, we have solved that $\hat{\beta}_i^{\MTL}=B^{\star}\left( U_r U_r(i) +\oo(1)\right)$. Inserting it into the definition of the test loss, we get that
	\begin{align*}
		\te(\hat{\beta}_t^{\MTL}) &= \bignorm{\Sigma^{1/2} \left((B^\star \hat{\cal W}^\top) (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat W_t - \beta_t \right) }^2
		+ \sigma^2  \hat W_t^{\top} (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat W_t \cdot \bigtr{\Sigma (X^{\top}X)^{-1}} \\
		&= \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} U_r U_r(t)-\beta_t}}^2 + \oo\left(\|B^\star\|^2\right) + \sigma^2\norm{U_r(t)}^2 \bigtr{\Sigma (X^{\top}X)^{-1}} \cdot (1+\oo(1))\\
		&= \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} U_r U_r(t)-\beta_t}}^2 + \frac{\sigma^2}{\rho-1}\norm{U_r(t)}^2 + \oo \left( \|B^\star\|^2 + \sigma^2\right),
	\end{align*}
	with high probability, where we used Lemma \ref{lem_minv} in the last step. Similar, by Lemma \ref{lem_minv} we have
	$$\te(\hat{\beta}_t^{\STL})=\frac{\sigma^2}{\rho-1} \cdot \left( 1+\oo(1)\right).$$
Combining the above two estimates, we conclude the proof.
\end{proof}
%From the above we can obtain three conceptual insights that are consistent with Section \ref{sec_denoise} and \ref{sec_insight}.
%\begin{itemize}
%	\item The de-noising effect of multi-task learning.
%	\item Multi-task training vs single-task training can be either positive or negative.
%	\item Transfer learning is better than the other two. And the improvement over multi-task training increases as the model distances become larger.
%\end{itemize}

>>>>

\begin{proof}[Proof of Theorem \ref{thm_many_tasks}]
In this setting, we need to study the following loss function:
\begin{align}
	f(B; W_1, \dots, W_t) = \sum_{i=1}^t \bignorm{X B W_i - Y_i}^2,\label{eq_mtl_same_cov}
\end{align}
for $W_1, W_2, \dots, W_t \in \R^r$. Then it is not hard to derive a closed form solution for $B$ as
	\begin{align*}
		\hat{B}(W_1, \dots, W_t) &= (X^{\top}X)^{-1} X^{\top} \bigbrace{\sum_{i=1}^t Y_i W_i^{\top}} (\cal W  \cal W^{\top})^{-1} \\
		&= (B^\star \cal W ^{\top}) (\cal W \cal W ^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cal W \cal W^{\top})^{-1},
	\end{align*}
	where $\cal W \in\real^{r\times t}$ is a matrix defined as $\cal W=[W_1, W_2, \dots, W_t]$.
	%Now we switch $\hat{B}$ back into equation \eqref{eq_mtl_same_cov} to
Then as in \eqref{approxvalid}, we pick $N$ independent samples of the training set for each task with $N\ge n^{1-\e_0}$, and use the concentration result, Lemma \ref{largedeviation}, to get the validation loss as 
\be\label{eq_multival}g(\cal W)=  N\left[\val(\cal W) + t  \sigma^2 \right]\cdot \left( 1+\OO(p^{-(1-\e_0)/2+\e})\right).\ee
Here $\val(\cal W)$ is defined as
%it remains to consider minimizing the validation loss
	$$\val( \cal W):=\exarg{\varepsilon_j, \forall 1\le j\le t}{ \sum_{i=1}^t \bignorm{\Sigma^{1/2}( \hat B W_i - \beta_i)}^2} =  \delta_{\bias}(\cal W) + \delta_{\vari} ,$$
where the model shift bias term $\delta_{\bias}(\cal W) $ is given by
	\begin{align*}
		\delta_{\bias}(\cal W) :=\sum_{i=1}^t  \bignorm{\Sigma^{1/2}\bigbrace{(B^\star \cal W^\top) (\cal W\cal W^{\top})^{-1} W_i - \beta_i}}^2,
	\end{align*}
	and the variance term $\delta_{\vari} $ can be calculated as
	\begin{align*}
		\delta_{\vari} := \sigma^2 \cdot \bigtr{\Sigma (X^{\top}X)^{-1}}.
	\end{align*}
It suffices to minimize $\delta_{\bias}(\cal W)$ over $\cal W$, since both $t \sigma^2$ and $\delta_{\vari}$ do not depend on $\cal W$.

We denote $Q := \cal W^{\top} (\cal W\cal W^{\top})^{-1} \cal W \in\real^{t\times t}$, whose $(i,j)$-th entry is given by $W_i^{\top} (\cal W\cal W^{\top})^{-1} W_j$.
	%Let $B^{\star} = [\beta_1, \beta_2, \dots, \beta_k] \in\real^{p \times k}$ denote the true model parameters.
	Now we can write $\delta_{\bias}(\cal W)$ succinctly as
	\begin{align*}
		\delta_{\bias}(Q) \equiv \delta_{\bias}(\cal W) = \bignormFro{\Sigma^{1/2}B^{\star}  \bigbrace{Q -\id}}^2 .
	\end{align*}
	From this equation we can solve the minimizer explicitly as $Q_0=U_{r}U_r^{\top}$. On the other hand, let $\hat {\cal W}$ be the minimizer of $g$, and denote $\hat Q:= \hat{\cal W}^{\top} (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat{\cal W} $. We claim that $\hat Q$ satisfies
	\be\label{Q-Q}\|Q_0^{-1}\hat Q - \id\|_F = \oo(1) \quad \text{w.h.p.}\ee
	In fact, if \eqref{Q-Q} does not hold, then using the condition $\lambda_{\min}((B^{\star})^\top\Sigma B^{\star})\gtrsim \sigma^2$ and that $\delta_{\vari}(\cal W)=\OO(\sigma^2)$ by  Lemma \ref{lem_minv}, we obtain that
	$$   \val(\hat {Q}) + t  \sigma^2 > (\val( Q_0) + t \sigma^2 )\cdot (1+\oo(1)) \ \Rightarrow \ g( \hat{\cal W})>g( {\cal W}_0),$$
	where $\cal W_0\in \R^{r\times t}$ is a matrix such that  $ \cal W_0^{\top} (\cal W_0\cal W_0^{\top})^{-1} \cal W_0=Q_0$. Hence  $\hat {\cal W}$ is not a minimizer, which leads to a contradiction.

	In sum, we have solved that $\hat{\beta}_i^{\MTL}=B^{\star}\left( U_r v_i +\oo(1)\right)$. Inserting it into the definition of the test loss, we get that with high probability,
	\begin{align*}
		\te(\hat{\beta}_t^{\MTL}) &= \bignorm{\Sigma^{1/2} \left((B^\star \hat{\cal W}^\top) (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat W_t - \beta_t \right) }^2
		+ \sigma^2  \hat W_t^{\top} (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat W_t \cdot \bigtr{\Sigma (X^{\top}X)^{-1}} \\
		&= \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} U_r v_t-\beta_t}}^2 + \oo\left(\|B^\star\|^2\right) + \sigma^2\norm{v_t}^2 \bigtr{\Sigma (X^{\top}X)^{-1}} \cdot (1+\oo(1))\\
		&= \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} U_r v_t-\beta_t}}^2 + \frac{\sigma^2}{\rho-1}\norm{v_t}^2 + \oo \left( \|B^\star\|^2 + \sigma^2\right),
	\end{align*}
 where we used Lemma \ref{lem_minv} in the last step. On the other hand, by Lemma \ref{lem_minv} we have
	$$\te(\hat{\beta}_t^{\STL})=\frac{\sigma^2}{\rho-1} \cdot \left( 1+\oo(1)\right).$$
Combining the above two estimates, we conclude the proof.
\end{proof}
%From the above we can obtain three conceptual insights that are consistent with Section \ref{sec_denoise} and \ref{sec_insight}.
%\begin{itemize}
%	\item The de-noising effect of multi-task learning.
%	\item Multi-task training vs single-task training can be either positive or negative.
%	\item Transfer learning is better than the other two. And the improvement over multi-task training increases as the model distances become larger.
%\end{itemize}
